Predictive Models to Support Quoting of Fixed Fee Consulting Projects

### 2016-12-21 finalwriteup/Consulting Profitability Paper/Consulting Profitability Paper.Rmd
new title

---
title: Predictive Models to Support Quoting of Fixed Fee Consulting Projects
author:
  - name: Amy Cook, Paul Wu, Kerrie Mengersen
    email: a21.cook@qut.edu.au
    affiliation: Queensland University of Technology
    footnote: Corresponding Author
address:
  - code: Queensland University of Technology
    address: School of Mathematical Sciences, George Street, Brisbane, QLD, 4000
abstract: |
  A concise abstract is required. limit to 250 words. clearly state purpose of research, principal resutls and major conclusions. no refs

  Engaging in loss making jobs for fixed fees is a major problem in consulting, particularly in the competitive construction industry. This thesis investigates whether machine learning techniques applied to a company's passively collected internal data could help avoid loss making jobs or help tactfully choose when to enforce stricter contracts. It was found that in a specific decision framework, a case study's profits could be improved 9% by declining approximately 4% of projects. Alternative decision frameworks are also proposed and evaluated. Algorithmic methods such as Logistic Regression, Random Forests, Boosted Trees, Naive Bayes, and Bayesian Networks were applied as well as blended combinations of these methods. A decision scenario which rejected projects above a sequence of tested thresholds was run in order to find the optimal threshold for profit improvements. The blended Logistic Regression model outperformed other methods and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. The findings from this research have the potential to assist managers in reducing losses by highlighting risky projects and guiding project-based changes to fee structures.
keywords: |
  consulting; machine learning; profitability; predictive model; construction industry; data mining
abbreviations: 
  - name: Customer Relationship Management
    footnote: Abbreviations

@@

From the profit curves, the optimal probability thresholds are where the profit curves peak, and peaks above 100% indicate the model assisted profits outperformed historical profits. The two simple blended models clearly outperformed all other models with higher profit ratios as well as tighter confidence intervals. The simple Logistic Regression blend performed best with the highest mean profit ratio of 109% and a standard deviation of 1.28%. This means that if all jobs above the probability threshold 0.6 were rejected, the profits would increase on average by 9% in comparison to historical profits. Only 4.3% of projects were rejected in this case.

@@

>>>>

---
title: Predictive Framework to Support Managerial Decisions around Fixed Fee Consulting Projects
author:
  - name: Amy Cook, Paul Wu, Kerrie Mengersen
    email: a21.cook@qut.edu.au
    affiliation: Queensland University of Technology
    footnote: Corresponding Author
address:
  - code: Queensland University of Technology
    address: School of Mathematical Sciences, George Street, Brisbane, QLD, 4000
abstract: |
  A concise abstract is required. limit to 250 words. clearly state purpose of research, principal resutls and major conclusions. no refs

  Fixed fee projects that run overbudget are a major problem for consulting businesses, particularly in competitive industries such as the construction industry. This thesis investigates whether machine learning algorithms trained on a company's passively collected internal data could predict whether a job would be profitable or loss making. Logistic Regression, Random Forests, Boosted Trees, and Naive Bayes were applied as well as blended combinations of these methods. The aim of these models is to help managers determine when to enforce certain contract structures or when to not engage in a project in order to avoid losses. Therefore this paper also examines whether the predictive algorithm improves a case study's overall profits if implementing the algorithm within a simple managerial decision framework. The framework was to reject all projects predicted as loss making, where loss making projects were defined as projects with a probability of being loss making above a certain threshold. The Logistic Regression model that blended predictive output from a Logistic Regression, Random Forest, and Boosted Tree models performed best and produced a 95% confidence interval of 6.5 - 11.5% profit improvements. This improvement was achieved by rejecting approximately 4% of projects that were deemed high risk. The findings from this research have the potential to assist managers in reducing consulting project losses by highlighting which projects to either avoid or accept given a fee structure that minimises financial risk.

keywords: |
  consulting; machine learning; profitability; predictive model; construction industry; data mining
abbreviations: 
  - name: Customer Relationship Management
    footnote: Abbreviations

@@

From the profit curves, the optimal probability thresholds are where the profit curves peak, and peaks above 100% indicate the model-assisted profits outperformed historical profits. The two simple blended models clearly outperformed all other models with higher profit ratios as well as tighter confidence intervals. The simple Logistic Regression blend performed best with the highest mean profit ratio of 109% and a standard deviation of 1.28%. This means that if all jobs above the probability threshold 0.6 were rejected, the profits would increase on average by 9% in comparison to historical profits. Only 4.3% of projects were rejected in this case.

| Optimal Threshold | Method               | Mean profit ratio  (%) | SD profit ratio  (%) | Lower bound 95% CI (%) | Upper bound 95% CI (%) |
|-----------|----------------------|------------------------|----------------------|------------------------|------------------------|
| 0.60      | Simple LR blend      | 109.0                  | 1.28                 | 106.5                  | 111.5                  |
| 0.55      | Simple average blend | 108.7                  | 1.56                 | 105.6                  | 111.8                  |
| 0.55      | Simple BT blend      | 105.6                  | 3.21                 | 99.3                   | 111.9                  |
| 0.50      | Complex BT blend     | 105.4                  | 3.23                 | 99.1                   | 111.7                  |
| 0.60      | Individual RF        | 103.2                  | 2.58                 | 98.1                   | 108.3                  |
| 0.60      | Individual BT        | 103.2                  | 2.77                 | 97.8                   | 108.6                  |
| 0.75      | Individual LR        | 102.5                  | 2.93                 | 96.8                   | 108.2                  |
| 0.65      | Complex RF blend     | 101.3                  | 2.72                 | 96.0                   | 106.6                  |
| 0.65      | Complex LR blend     | 101.0                  | 1.18                 | 98.7                   | 103.3                  |

@@

Table Captions
===============
1. Values and statistics describing each individual model and blended model's profit curves at the optimal position in the curve.

### 2016-12-21 finalwriteup/Consulting Profitability Paper/Consulting_Profitability_Paper.tex
new title

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Predictive Models to Support Quoting of Fixed Fee Consulting Projects},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  Engaging in loss making jobs for fixed fees is a major problem in
  consulting, particularly in the competitive construction industry. This
  thesis investigates whether machine learning techniques applied to a
  company's passively collected internal data could help avoid loss making
  jobs or help tactfully choose when to enforce stricter contracts. It was
  found that in a specific decision framework, a case study's profits
  could be improved 9\% by declining approximately 4\% of projects.
  Alternative decision frameworks are also proposed and evaluated.
  Algorithmic methods such as Logistic Regression, Random Forests, Boosted
  Trees, Naive Bayes, and Bayesian Networks were applied as well as
  blended combinations of these methods. A decision scenario which
  rejected projects above a sequence of tested thresholds was run in order
  to find the optimal threshold for profit improvements. The blended
  Logistic Regression model outperformed other methods and produced a 95\%
  confidence interval of 6.5 - 11.5\% profit improvements. The findings
  from this research have the potential to assist managers in reducing
  losses by highlighting risky projects and guiding project-based changes
  to fee structures.
  \end{abstract}
   \begin{keyword} consulting; machine learning; profitability; predictive model;
construction industry; data mining \sep \end{keyword}
 \end{frontmatter}

@@

From the profit curves, the optimal probability thresholds are where the
profit curves peak, and peaks above 100\% indicate the model assisted
profits outperformed historical profits. The two simple blended models
clearly outperformed all other models with higher profit ratios as well
as tighter confidence intervals. The simple Logistic Regression blend
performed best with the highest mean profit ratio of 109\% and a
standard deviation of 1.28\%. This means that if all jobs above the
probability threshold 0.6 were rejected, the profits would increase on
average by 9\% in comparison to historical profits. Only 4.3\% of
projects were rejected in this case.

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Predictive Models to Support Quoting of Fixed Fee Consulting Projects},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  Fixed fee projects that run overbudget are a major problem for
  consulting businesses, particularly in competitive industries such as
  the construction industry. This thesis investigates whether machine
  learning algorithms trained on a company's passively collected internal
  data could predict whether a job would be profitable or loss making.
  Logistic Regression, Random Forests, Boosted Trees, and Naive Bayes were
  applied as well as blended combinations of these methods. The aim of
  these models is to help managers determine when to enforce certain
  contract structures or when to not engage in a project in order to avoid
  losses. Therefore this paper also examines whether the predictive
  algorithm improves a case study's overall profits if implementing the
  algorithm within a simple managerial decision framework. The framework
  was to reject all projects predicted as loss making, where loss making
  projects were defined as projects with a probability of being loss
  making above a certain threshold. The Logistic Regression model that
  blended predictive output from a Logistic Regression, Random Forest, and
  Boosted Tree models performed best and produced a 95\% confidence
  interval of 6.5 - 11.5\% profit improvements. This improvement was
  achieved by rejecting approximately 4\% of projects that were deemed
  high risk. The findings from this research have the potential to assist
  managers in reducing consulting project losses by highlighting which
  projects to either avoid or accept given a fee structure that minimises
  financial risk.
  \end{abstract}
   \begin{keyword} consulting; machine learning; profitability; predictive model;
construction industry; data mining \sep \end{keyword}
 \end{frontmatter}

@@

From the profit curves, the optimal probability thresholds are where the
profit curves peak, and peaks above 100\% indicate the model-assisted
profits outperformed historical profits. The two simple blended models
clearly outperformed all other models with higher profit ratios as well
as tighter confidence intervals. The simple Logistic Regression blend
performed best with the highest mean profit ratio of 109\% and a
standard deviation of 1.28\%. This means that if all jobs above the
probability threshold 0.6 were rejected, the profits would increase on
average by 9\% in comparison to historical profits. Only 4.3\% of
projects were rejected in this case.

\begin{longtable}[]{@{}llllll@{}}
\toprule
\begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
Optimal Threshold\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright\strut
Method\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Mean profit ratio (\%)\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright\strut
SD profit ratio (\%)\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Lower bound 95\% CI (\%)\strut
\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Upper bound 95\% CI (\%)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.60\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Simple LR blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
109.0\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
1.28\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
106.5\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
111.5\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.55\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Simple average blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
108.7\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
1.56\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
105.6\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
111.8\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.55\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Simple BT blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
105.6\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
3.21\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
99.3\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
111.9\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.50\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Complex BT blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
105.4\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
3.23\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
99.1\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
111.7\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.60\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Individual RF\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
103.2\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
2.58\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
98.1\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
108.3\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.60\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Individual BT\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
103.2\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
2.77\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
97.8\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
108.6\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.75\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Individual LR\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
102.5\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
2.93\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
96.8\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
108.2\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.65\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Complex RF blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
101.3\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
2.72\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
96.0\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
106.6\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
0.65\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
Complex LR blend\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
101.0\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright\strut
1.18\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
98.7\strut
\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
103.3\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\section{Table Captions}\label{table-captions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Values and statistics describing each individual model and blended
  model's profit curves at the optimal position in the curve.
\end{enumerate}
